---
title: "Introducing qualpalr"
author: "Johan Larsson"
date: 2016-10-15
categories: ["R"]
tags: ["qualpalr", "color-science"]
bibliography: [../../static/library.bib]
header:
  image: "headers/france.svg"
summary: "Let me introduce qualpalr: an R package that generates qualitative 
          color palettes with distinct colors using color difference
          algorithms."
---

With the advent of [colorbrewer](http://colorbrewer2.org/) there now exists good
options to generate color palettes for sequential, diverging, and qualitative 
data. In R, these palettes can be accessed via the popular
[RColorBrewer](https://cran.r-project.org/package=RColorBrewer) package. Those
palettes, however, are limited to a fixed number of colors. This isn't much of 
a problem for sequential of diverging data since we can interpolate colors to 
any range we desire:

```{r}
set.seed(1)
pal <- RColorBrewer::brewer.pal(4, "PuBuGn")
color_ramp <- colorRampPalette(pal, space = "Lab")
```

There is not, however, an analogue for qualitative color palettes that will get
you beyond the limits of 8--12 colors of colorbrewer's qualitative color 
palettes. 

There is also no customization in colorbrewer. Other R packages, such as 
[colorspace](https://cran.r-project.org/package=colorspace) offer this, but 
they are primarily adapted to sequential and diverging data -- not qualitative
data.

This is where qualpalr comes in. qualpalr provides the user with a convenient
way of generating distinct qualitative color palettes, primarily for use in 
R graphics. Given `n` (the number of colors to generate), along with a subset
in the [hsl color space](https://en.wikipedia.org/wiki/HSL_and_HSV) (a 
cylindrical representation of the RGB color space) `qualpalr` attempts to find
the `n` colors in the provided color subspace that 
*maximize the smallest pairwise color difference*. This is done by projecting 
the color subset from the HSL color space to the DIN99d space. DIN99d is 
(approximately) perceptually uniform, that is, the euclidean distance between
two colors in the space is proportional to their perceived difference.

## Examples

`qualpalr` relies on one basic function, `qualpal()`, which takes as its input
`n` (the number of colors to generate) and `colorspace`, which can be either

* a list of numeric vectors `h` (hue from -360 to 360), `s` (saturation from 0 
to 1), and `l` (lightness from 0 to 1), all of length 2, specifying a min and 
max.
* a character vector specifying one of the predefined color subspaces, which at 
the time of writing are *pretty*, *pretty_dark*, *rainbow*, and *pastels*.

```{r basic_usage}
library(qualpalr)
pal <- qualpal(n = 5, list(h = c(0, 360), s = c(0.4, 0.6), l = c(0.5, 0.85)))

# Adapt the color space to deuteranopia
pal <- qualpal(n = 5, colorspace = "pretty", cvd = "deutan")
```

The resulting object, `pal`, is a list with several color tables and a distance 
matrix based on the din99d color difference formula.

```{r qualpal_list}
pal
```

Methods for `pairs` and `plot` have been written for `qualpal` objects to help
visualize the results.

```{r mds, fig.width = 5, figh.height = 5, fig.cap = "Multi-dimensional scaling plots, displaying the difference in colors through their distances."}
# Multidimensional scaling plot
plot(pal)
```

```{r pairs, fig.width = 5, fig.height = 5, fig.cap = "Scatterplot matrix of colors in the DIN99d color space."}
# Pairs plot in the din99d color space
pairs(pal, colorspace = "DIN99d")
```

The colors are normally used in R by fetching the `hex` attribute of the 
palette.

```{r map, fig.width=5, fig.height=5, fig.cap = "A map of france, using colors from qualpalr."}
library(maps)
map("france", fill = TRUE, col = pal$hex)
```

## Details

`qualpal` begins by generating a point cloud out of the HSL color subspace 
provided by the user, using a quasi-random torus sequence from 
[randtoolbox](https://cran.r-project.org/package=randtoolbox). Here is the 
color subset in HSL with settings 
`h = c(-200, 120), s = c(0.3, 0.8), l = c(0.4, 0.9)`.

```{r hsl, echo = FALSE, fig.cap = "A cloud of colors in HSL space."}
options(rgl.useNULL = TRUE)
library(rgl)

# Set up color subspace as in qualpal()
h = c(-200, 120)
s = c(0.3, 0.8)
l = c(0.4, 0.9)

# Generate a quasi-random tour sequence
rnd <- randtoolbox::sobol(n = 1000, dim = 3, scrambling = 2)

# Convert random sequence to specified color space
scale_runif <- function(x, new_min, new_max) {
  (new_max - new_min) * (x - 1) + new_max
}

rnd[, 2] <- sqrt(rnd[, 2])
H <- scale_runif(rnd[, 1], min(h), max(h))
S <- scale_runif(rnd[, 2], min(s), max(s))
L <- scale_runif(rnd[, 3], min(l), max(l))

HSL <- cbind(H, S, L)
HSL[HSL[, 1] < 0, 1] <- HSL[HSL[, 1] < 0, 1] + 360

# Set up variables to produce the HSL cylinder
x <- S * cos(H * pi / 180)
y <- S * sin(H * pi / 180)

# Convert to RGB to enable coloring
RGB <- qualpalr:::HSL_RGB(HSL)

plot3d(cbind(x, y, L), col = grDevices::rgb(RGB), main = "HSL")
rglwidget()
```

The function then proceeds by projecting these colors into the sRGB space.

```{r rgb, echo = FALSE, fig.cap = "A 3D cloud of colors in sRGB space."}
plot3d(RGB, col = grDevices::rgb(RGB), main = "RGB")
rglwidget()
```

It then continues by projecting the colors, first into the XYZ space, then
CIELab (not shown here), and then finally the DIN99d space.

```{r din, echo = FALSE, fig.cap = "A 3D cloud of colors in DIN99d space."}
XYZ    <- qualpalr:::sRGB_XYZ(RGB)
DIN99d <- qualpalr:::XYZ_DIN99d(XYZ)

plot3d(DIN99d, col = rgb(RGB), main = "DIN99d")
rglwidget()
```

The DIN99d color space [@cui_uniform_2002] is a euclidean, perceptually uniform 
color space. This means that the difference between two colors is equal to the
euclidean distance between them. We take advantage of this by computing a 
distance matrix on all the colors in the subset, finding their pairwise color 
differences. We then apply a power transformation [@huang_power_2015] to fine 
tune these differences.

To select the `n` colors that the user wanted, we proceed greedily: first, we 
find the two most distant points, then we find the third point that maximizes 
the minimum distance to the previously selected points. This is repeated until
`n` points are selected. These points are then returned to the user; below is 
an example using `n = 5.`

```{r choice, echo = FALSE, fig.cap = "The selected colors (in DIN99d space)."}
pal <- qualpal(5, list(h = c(-200, 120), s = c(0.3, 0.8), l = c(0.4, 0.9)))
plot3d(DIN99d, col = rgb(RGB), main = "DIN99d", alpha = 0.1)
plot3d(pal$DIN99d, col = pal$hex, main = "DIN99d", add = TRUE, size = 5)
rglwidget()
```

### Color specifications

At the time of writing, qualpalr only works in the sRGB color space with
the CIE Standard Illuminant D65 reference white. 

## Future directions

The greedy search to find distinct colors is crude. Particularly when searching
for few colors, the greedy algorithm will lead to sub-optimal results. Other
solutions to finding points that maximize the smallest pairwise distance 
among them are welcome.

## Thanks

[Bruce Lindbloom's webpage](http://www.brucelindbloom.com/) has
been instrumental in making qualpalr. Also thanks to
[i want hue](http://tools.medialab.sciences-po.fr/iwanthue/), which inspired me
to make qualpalr.

## References
