---
title: "Introducing eulerr"
author: "Johan Larsson"
date: 2016-10-19
categories: ["R"]
tags: ["eulerr", "euler-diagrams"]
bibliography: [../../library.bib]
---

**eulerr** is an [R](https://www.r-project.org) package that generates
area-proportional euler diagrams to display set
relationships (intersections, unions, and disjoints) with circles.
[Euler diagrams](https://en.wikipedia.org/wiki/Euler_diagram) are Venn
diagrams without the requirement that all set interactions be present (whether
they are empty or not). That is, depending on input, eulerr will sometimes
produce Venn diagrams but sometimes not.

## Background

R features a number of packages that produce euler and/or venn diagrams;
some of the more prominent ones (on CRAN) are

* [eVenn](https://cran.r-project.org/package=eVenn),
* [VennDiagram](https://cran.r-project.org/package=VennDiagram),
* [venn](https://cran.r-project.org/package=venn),
* [colorfulVennPlot](https://cran.r-project.org/package=colorfulVennPlot), and
* [venneuler](https://cran.r-project.org/package=venneuler).

The last of these (venneuler) serves as the primary inspiration for this package,
along with the refinements that Ben Fredrickson has presented on his
[blog](http://www.benfrederickson.com/) and made available in his javascript
[venn.js](https://github.com/benfred/venn.js).

venneuler, however, is written in java, preventing R users from 
browsing the source code (unless they are also literate in java) or
contributing. Furthermore, venneuler is known to produce imperfect output for
set configurations that have perfect solutions. Consider,
for instance, the following example in which the intersection between `A` and
`B` is unwanted.

```{r venneuler-failure, fig.height=5, fight.width=5, fig.cap='venneuler plot with unwanted overlap.'}
library(venneuler, quietly = TRUE)
venn_fit <- venneuler(c(A = 75, B = 50, "A&B" = 0))
plot(venn_fit)
```

## Enter eulerr

eulerr is based on the improvements to **venneuler** that Ben Fredrickson
introcued with **venn.js** but has been coded from scratch, uses different
optimizers, and returns the residuals and stress statistic that venneuler
features. 

### Input

Currently, it is possible to provide input to `eulerr` as either

* a named numeric vector or
* a matrix of logicals with columns representing sets and rows the set
relationships for each observation.

```{r input}
library(eulerr)

# Input in the form of a named numeric vector
fit1 <- euler(c("A" = 25, "B" = 5, "C" = 5,
                "A&B" = 5, "A&C" = 5, "B&C" = 3,
                "A&B&C" = 3))

# Input as a matrix of logicals
set.seed(1)
mat <-
  cbind(
    A = sample(c(TRUE, TRUE, FALSE), size = 50, replace = TRUE),
    B = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
    C = sample(c(TRUE, FALSE, FALSE, FALSE), size = 50, replace = TRUE)
  )
fit2 <- euler(mat)
```

### Fit

We inspect our results by printing the eulerr object

```{r fit1_print}
fit2
```

or directly access and plot the residuals.

```{r residual-plot, fig.cap='Residuals for the eulerr fit.', fig.height = 5, fig.width = 5}
# Cleveland dot plot of the residuals
dotchart(resid(fit2))
abline(v = 0, lty = 3)
```

This shows us that the `A&B&C` intersection is somewhat overrepresented in
`fit2`. Given that these residuals are on the scale of the original
values, however, the residuals are arguably of little concern.

For an overall measure of the fit of the solution, we use the same stress
statistic that Leland Wilkinson presented in his academic paper on venneuler 
(@wilkinson_exact_2012), which is given by the sums of squared residuals divided
by the total sums of squares:
$$\frac{\sum_{i=1}^n (f_i - y_i)^2}{\sum_{i=1}^n (y_i - \bar{y})^2}$$

We fetch it from the `stress` attribute of the `eulerr` object.

```{r stress}
fit2$stress
```

We can now be confident that eulerr provides a reasonable representation of
our input. Were it otherwise, we would do best to stop here and look for another
way to visualize our data. (I suggest the excellent
[UpSetR](https://cran.r-project.org/package=UpSetR) package.)

### Plotting

No we get to the fun part: plotting our diagram. This is easy, as well as
highly customizable, with eulerr.

```{r eulerr-plot, fig.width=3.5, fig.height=3.5, fig.show='hold', fig.cap='eulerr plots can be modified in every possible way.'}
plot(fit2)

# Change fill colors, border type (remove) and fontface.
plot(fit2,
     fill = list(col = c("dodgerblue4", "darkgoldenrod1", "cornsilk4")),
     edges = list(lty = 1:3),
     labels = list(font = 2))
```

eulerr's default color palette is taken from
[qualpalr](https://cran.r-project.org/package=qualpalr) -- another
package that I have developed -- which uses color difference algorithms to 
generate distinct qualitative color palettes.

## Details

Details of the implementation will be left for a future vignette but almost
completely resemble the approach documented
[here](http://www.benfrederickson.com/better-venn-diagrams/).

## Thanks

eulerr would not be possible without Ben Fredrickson's work on
[venn.js](http://www.benfrederickson.com) or Leland Wilkinson's
[venneuler](https://cran.r-project.org/package=venneuler).

## References
