---
title: Polygon labeling with polylabelr
slug: polygon-labeling-with-polylabelr
categories:
  - R
tags:
  - euler diagrams
  - polygon
  - labeling
  - R
  - eulerr
  - polylabelr
  - geometry
excerpt: Using polylabelr to find visual centers of polygons and
  label Euler diagrams
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  dev.args = list(pointsize = 8),
  crop = TRUE,
  fig.align = "center",
  fig.width = 3.5,
  fig.height = 2.9
)

options(digits = 3)
```

## Labeling Euler diagram overlaps

The purpose of my R package [eulerr](https://github.com/jolars/eulerr)
is to fit and *visualize* Euler diagrams. Besides the various intricacies 
involved in fitting the diagrams, there are many interesting
problems involved in their visualization. One of these is the labeling of the 
overlaps.

Naturally, simply positioning the labels at the shapes' centers
fails more often than not. Nevertheless, this stategy is employed by
**venneuler**, for instance, and the plots usually demand
manual tuning.

```{r, fig.cap = "A plot from venneuler with suboptimal label placements."}
# an example set combination
s <- c("SE" = 13, "Treat" = 28, "Anti-CCP" = 101, "DAS28" = 91,
       "SE&Treat" = 1, "SE&DAS28" = 14, "Treat&Anti-CCP" = 6,
       "SE&Anti-CCP&DAS28" = 1)

library(venneuler, quietly = TRUE)
fit_venneuler <- venneuler(s)
plot(fit_venneuler)
```

Up til now, I solved this in **eulerr** by, for each overlap,
filling one of the involved shapes (circles or ellipses) with points
and then numerically optimizing the location of the point using a
Nelder--Mead optimizer. However, given that the solution to
finding the distance between a point and an ellipse---at least one that
is rotated---itself requires a numerical solution [@eberly2013], this procedure
turned out to be quite inefficient.

## The promise of polygons

R has powerful functionality for plotting in general, but lacks 
capabilities for drawing ellipses using curves. High-resolution 
polygons are thankfully a readily available remedy for this and have
since several version back been used also in **eulerr**. 

The upside of using polygons, however, are that they are usually
much easier, even if sometimes inefficient,
to work with. For instance, they make constructing separate shapes
for each overlap a breeze using the polyclip package [@johnson2018].

And because basically all shapes in digital maps are polygons,
there happens to exist a multitude of other useful tools to deal with
a wide variety of tasks related to polygons. One of these turned out
to be precisely what I needed: polylabel [@mapbox2018a] from the Mapbox suite.
Because the details of the library
[have already been explained elsewhere](https://blog.mapbox.com/a-new-algorithm-for-finding-a-visual-center-of-a-polygon-7c77e6492fbc)
I will spare you the details, but briefly put it uses quadtree
binning to divide the polygon into square bins, pruning away dead-ends.
It is inefficient and will, according to the authors, find
a point that is "guaranteed to be a global optimum within the given precision".

Because it appeared to be such a valuable tool for R users, I decided
to create a wrapper for the c++ header for polylabel and bundle it as a package for R users.

```{r, fig.cap = "Locating poles of inaccessibility with polylabel."}
# install.packages("polylabelr")
library(polylabelr)

# a concave polygon with a hole
x <- c(0, 6, 3, 9, 10, 12, 4, 0, NA, 2, 5, 3)
y <- c(0, 0, 1, 3, 1, 5, 3, 0, NA, 1, 2, 2)

# locate the pole of inaccessibility
p <- poi(x, y, precision = 0.01)

plot.new()
plot.window(range(x, na.rm = TRUE), range(y, na.rm = TRUE), asp = 1)
polypath(x, y, col = "grey90", rule = "evenodd")
points(p, cex = 2, pch = 16)
```

The package [is availabe on cran](https://CRAN.R-project.org/package=polylabelr),
the source code is located at <https://github.com/jolars/polylabelr> and
is documented at <https://jolars.github.io/polylabelr/>.

## Euler diagrams

To come back around to where we started at, **polylabelr** has now been
employed in the development branch 
of **eulerr** where it is used to quickly and appropriately 
figure out locations for the labels of the diagram.

```{r, fig.cap = "An Euler diagram with appropriate label placement."}
library(eulerr)

plot(euler(s))
```

## References
